---
title: "Analyzing The Influence of Biological Factors vs. Lifestyle Habits on Stroke Victims"
author: "Group 4"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_document
---

Biological Factors

| | |
| --- | --- |
| gender | categorical |
| age | numeric |
| hypertension | categorical |
| heart_disease | categorical |
| avg_glucose_level |
| bmi | categorical |


Lifestyle Habits:

- ever_married
- work_type
- Residence_type
- smoking_status 

```{r setup}
set.seed(10);

library(ggplot2);
library(dplyr);
library(knitr);
library(car);
library(ROCR);
library(sjPlot);
library(ggcorrplot);

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r colors}
UCLA_colors <- list(
    ucla_blue = '#2774AE',
    ucla_gold = '#FFD100',
    darkest_blue = '#003B5C',
    darker_blue = '#005587',
    lighter_blue = '#8BB8E8',
    lightest_blue = '#DAEBFE',
    darkest_gold = '#FFB81C',
    darker_gold = '#FFC72C'
);
```


```{r data-prep}
load_stroke_data <- function(path) {
    read.csv(path, header = T, na.strings = "N/A");
}

prep_stroke_data <- function(stroke_data) {
    stroke_data <- na.omit(stroke_data);
    stroke_data <- stroke_data[-which(stroke_data$gender == "Other"), ];
    stroke_data$smoking_status <- factor(
        stroke_data$smoking_status,
        levels = c(
            'never smoked',
            'formerly smoked',
            'smokes'
        )
    );
    stroke_data$work_type <- factor(
        stroke_data$work_type,
        levels = c(
            'Never_worked',
            'Private',
            'Self-employed',
            'Govt_job',
            'children'
        )
    );
    
    return(stroke_data);
}

split_stroke_data <- function(stroke_data, training_proportion) {
    n <- nrow(stroke_data);
    num_training <- n * training_proportion;
    training_i <- sample.int(n, num_training);
    
    list(
        training = stroke_data[training_i, ],
        test = stroke_data[-training_i, ]
    );
}

stroke <- load_stroke_data("healthcare-dataset-stroke-data.csv") |> prep_stroke_data();
stroke <- split_stroke_data(stroke, training_proportion = 0.8);
```


# EDA
```{r eda}
ggplot(stroke$training, aes(x = age)) +
    geom_histogram() + ggtitle("Age");
    
ggplot(stroke$training, aes(x = avg_glucose_level)) +
    geom_histogram() + ggtitle("Glucose Level");
  
ggplot(stroke$training, aes(x = bmi)) +
    geom_histogram() + ggtitle("BMI");
```

## Correlation
```{r correlation}
# we should do this for all variables by converting to dummy, I selected only numeric ones for now
stroke_correlation <- cor(stroke$training |> subset(select = c(age, avg_glucose_level, bmi)));

ggcorrplot(
    stroke_correlation,
    method = 'circle',
    ggtheme = ggplot2::theme_gray,
    colors = c(UCLA_colors$ucla_blue, 'white', UCLA_colors$ucla_gold)
);
```

```{r homoscedasticity}
####### Equality of Variance #######
homoscedasticity_test <- function(stroke_data) {
    predictors <- colnames(stroke_data |> subset(select = -c(stroke, id, age, avg_glucose_level, bmi)))
    p <- sapply(
        predictors,
        function(column) bartlett.test(stroke ~ stroke_data[, column], data = stroke_data)$p.value
    );
    
    data.frame(
        p_value = p,
        equality_of_variance = p < 0.05
    );
}

knitr::kable(homoscedasticity_test(stroke$training));
```

## Age Groups

```{r age-groups}
group_ages <- function(stroke_data, breaks) {
    stroke_data$age_group <- factor(cut(stroke_data$age, breaks = breaks));
    return(stroke_data);
}

age_breaks <- c(0, 40, 60, 85);
stroke$training <- group_ages(stroke$training, breaks = age_breaks);
stroke$test <- group_ages(stroke$test, breaks = age_breaks);
```

```{r plot-age}
ggplot(
    stroke$training |> arrange(stroke) |> count(age_group, stroke), 
    aes(x = age_group, y = n, fill = factor(stroke, levels = c(1, 0)))) + 
        geom_bar(stat = 'identity', position = 'stack') +
        scale_fill_manual(
            labels = c('Stroke', 'No Stroke'),
            values = c(UCLA_colors$lighter_blue, UCLA_colors$darkest_blue),
            name = ''
        ) + theme(
            legend.text = element_text(size=11),
            axis.text = element_text(size=9),
            axis.title = element_text(size=14)
        ) +
        xlab('Age Group (Years)') + ylab('Number of Patients');
```

## Average Glucose Level
```{r avg-glucose}
symbox_plot <- symbox(~avg_glucose_level, data = stroke$training);
```

```{r diabetes}
split_diabetes <- function(stroke_data, breakpoint) {
    labels <- c('Normal', 'High');
    group_assignments <- as.numeric(cut(
        stroke_data$avg_glucose_level,
        breaks = c(0, breakpoint, Inf),
        labels = FALSE
    ));
    stroke_data$diabetes <- factor(
        labels[group_assignments],
        levels = labels
    );
    return(stroke_data);
}

diabetes_threshold <- 125;
stroke$training <- split_diabetes(stroke$training, diabetes_threshold);
stroke$test <- split_diabetes(stroke$test, diabetes_threshold);
```

```{r plot-diabetes}
ggplot(stroke$training, aes(x = avg_glucose_level)) +
    geom_histogram(
        fill = UCLA_colors$lighter_blue,
        breaks = seq(55, 275, by = 5)
    ) +
    xlab('Average Glucose Level (mg/dL)') + ylab('Number of Patients');

ggplot(stroke$training, aes(x = diabetes, y = avg_glucose_level)) +
    geom_boxplot(fill = UCLA_colors$lighter_blue) +
    xlab('Blood Sugar Group') + ylab('Average Glucose Level');
```

## Hypertension
```{r hypertension-eda}
hypertension_proportions <- stroke$training |> group_by(hypertension, stroke) |>
  summarise(n = n()) |>
  mutate(freq = n / sum(n)) |> ungroup();

ggplot(
    hypertension_proportions,
    aes(x = hypertension, y = freq, fill = factor(stroke, levels = 1:0))
) + geom_bar(
    stat = 'identity'
) + xlab('Hypertension') + ylab('Proportion of Patients') +
    scale_x_continuous(breaks = 0:1, labels = c('No', 'Yes')) +
    scale_fill_manual(
            labels = c('Stroke', 'No Stroke'),
            values = c(UCLA_colors$lighter_blue, UCLA_colors$darkest_blue),
            name = ''
        );
```


## Colinearity
```{r vif}
full_vif <- as.data.frame(vif(glm(
    stroke ~ . - id - age_group - diabetes,
    data = stroke$training,
    family = "binomial"
)));
full_vif$variable <- rownames(full_vif);

ggplot(full_vif, aes(x = GVIF, y = variable)) + 
    geom_bar(stat = 'identity', fill = UCLA_colors$ucla_blue) +
    scale_x_continuous(breaks = 1:6) + 
    geom_vline(
        xintercept = 5,
        linetype = "dashed", 
        color = UCLA_colors$lighter_blue,
        size = 0.5
    ) + xlab('VIF') + ylab('Variable');
```

### Biological Variables
```{r eda-bio}
biological_model <- glm(
    stroke ~ gender + age + hypertension + heart_disease + avg_glucose_level + bmi,
    data = stroke$training,
    family = 'binomial'
);
summary(biological_model);

biological_age_group_model <- glm(
    stroke ~ gender + age_group + hypertension + heart_disease + diabetes + bmi,
    data = stroke$training,
    family = 'binomial'
);
summary(biological_age_group_model);
```
### Lifestyle Variables
```{R eda-life}
lifestyle_model <- glm(
    stroke ~ ever_married + work_type + Residence_type + smoking_status,
    data = stroke$training,
    family = 'binomial'
);

summary(lifestyle_model);
```

### Full Model
```{r full-model}
full_model <- glm(
    stroke ~ age_group + gender + diabetes + heart_disease + hypertension + bmi + 
        ever_married + work_type + Residence_type + smoking_status,
    data = stroke$training,
    family = 'binomial'
);

summary(full_model);
data.frame(coef = round(full_model$coefficients, digits = 3));
```

### Interactions
This model includes all significant features from each set, plus targeted interaction effects for lifestyle and socio-economic factors that we hypothesize might have a relationship with health factors.
```{r eda-interaction}
interaction_model <- glm(
    stroke ~ 
        hypertension * age_group + 
        bmi * age_group + 
        avg_glucose_level * age_group +
        avg_glucose_level * bmi +
        heart_disease * age_group +
        smoking_status +
        Residence_type + 
        ever_married,
    data = stroke$training,
    family = 'binomial'
);
summary(interaction_model);
```

### Restricted Model
```{r eda-restricted}
restricted_model <- glm(
    stroke ~ 
        age_group + hypertension + age_group : diabetes,
    data = stroke$training,
    family = 'binomial'
);
summary(restricted_model);
```

### Coefficients
```{r coeff}
knitr::kable(
    data.frame(
        coefficient = round(exp(restricted_model$coefficients), digits = 3)
    )
);
```

# Model Accuracy
```{r accuracy}
model_accuracy <- function(model_pred){
  accuracy <- mean(model_pred == stroke$training$stroke)
  return(paste("Accuracy =", round(accuracy, 4)))
}

roc <- function(model_pred){
  roc_pred <- prediction(model_pred, stroke$training$stroke)
  roc_perf <- performance(roc_pred, measure = "tpr", x.measure = "fpr")
  
  auc <- performance(roc_pred, measure = "auc")
  auc <- auc@y.values[[1]]
   
  # Plotting
  plot(roc_perf, main = "ROC CURVE", col = "blue", lwd = 1.5)
  legend(.6, .4, round(auc, 4), title = "AUC", cex = 1)
}

confusion_matrix <- function(model_pred) {
  table(model_pred, stroke$training$stroke)
}

log_pred <- predict(restricted_model, stroke$training, type = "response")
```


```{r}
roc(log_pred)
```
Using Training data to determine Threshold:

```{r}
log_pred_2 <- ifelse(log_pred >= 0.05, 1, 0)
model_accuracy(log_pred_2)
confusion_matrix(log_pred_2)

model_accuracy_test <- function(model_pred){
  accuracy <- mean(model_pred == stroke$test$stroke)
  return(paste("Accuracy =", round(accuracy, 4)))
}
confusion_matrix_test <- function(model_pred) {
  table(model_pred, stroke$test$stroke)
}
```



```{r}
log_pred_test <- predict(restricted_model, stroke$test, type = "response")
log_pred_class_test <- ifelse(log_pred_test >= 0.05, 1, 0)
model_accuracy_test(log_pred_class_test)
confusion_matrix_test(log_pred_class_test)
```

Log Odds, Confidence Intervals, and P_Values
```{r final_interpretation}
## P_value
summary(restricted_model);
## Log Odds
coef(summary(restricted_model))[, 1];
## Confidence Intervals
conf_int <- confint(restricted_model)[-1, ];
lower_bound <- conf_int[, 1];
upper_bound <- conf_int[, 2];
mean_points <- apply(conf_int, 1, mean);

#Confidence Interval Plot
plot_model(restricted_model, 
           transform = NULL, 
           colors = c(UCLA_colors$darkest_gold, UCLA_colors$darkest_blue)) + 
          theme_bw()
```
